Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-27 sections/lang.tex
cool
fixed

%What are values? (objects in the host language)
%What is t really? (

@@

    \begin{prooftree}
    \hypo{\langle val, (e, t) \rangle
    \rightarrow val'}
    \infer1[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \hypo{}
        \infer1[$Literal$]{\langle\finchliteral(val), (e, t)\rangle \rightarrow val}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{}
        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t)\rangle}{\rightarrow e(\finchvar(name))}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t)\rangle}{\rightarrow e(\finchindex(name))}}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \hypo{\langle head, s_1 \rangle \rightarrow s_2}
        \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow true}
        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow false}
        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(tns) = tns'}
    \hypo{\langle init, s \rangle \Rightarrow init'}
    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
    \hypo{\langle declare(tns', init', dims'...), t \rangle \rightarrow t'}
    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], t')}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{e(tns) = tns'}
    \hypo{\langle freeze(tns'), t \rangle \rightarrow t'}
    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], t')}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = read}
    \hypo{e(tns) = tns'}
    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
    \end{prooftree}

@@

\vspace{12pt}

\begin{prooftree}
    \infer0[$Run$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(run(body), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(body, j...)]), s\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{e(i) = i'}
    \hypo{\langle seek(i'), t \rangle \rightarrow t'}
    \infer2[$Lookup$]{\splitfrac{\langle E[\finchaccess(lookup(seek, body), j..., i)], (e, t)\rangle}{\rightarrow \langle E[\finchaccess(body, j...)], (e, t')\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle cond, t \rangle \Rightarrow true}
    \infer1[]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(head, i...)], s\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{\langle cond, t \rangle \Rightarrow false}
    \infer1[$Switch$]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(tail, i...)], s\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, extent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle \finchloop(i, ext, E[\finchaccess(head, j..., i)]), s\rangle \rightarrow s'}
    \infer1[$Sequence$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(head, tail), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(tail, j..., i)]), s'\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{\langle node, algebra \rangle \rightarrow node'}
    \infer1[$Simplify$]{\langle E[node], s\rangle \rightarrow \langle E[node'], s\rangle}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle seek(a), t \rangle \rightarrow t'}
    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t')\rangle}}
\end{prooftree}
\vspace{12pt}

\begin{prooftree}
    \hypo{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t) \rangle \rightarrow (e', t')}
    \hypo{\langle next, t' \rangle \rightarrow t''}
    \infer2[$StepperNext$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), (e', t'')\rangle}}
\end{prooftree}

    \caption{Looplet evaluation semantics. 
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the host
    language. This means that rules which modify $t$ are running in the host language. Note that $E$ is an evaluation context that applies anywhere in
    the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to
    hoist conditions and subranges out of loops. However, this also means we must specify
    the priority in which we apply looplet rules, which is as follows:
    $Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup$
    Many looplets, most notably the thunk looplet, introduce variables into the
    host language environment.  While variables introduced by a looplet may be
    modified by the looplet itself (steppers often increment some state
    variables), we forbid child looplets from modifying any state variables that
    they didn't introduce. This allows us to treat the \finchvalue node as a
    constant.
    %
    Note: the $Simplify$ rule references $algebra$, which is our variable
    defining a set of straightforward simplification rules. These rules include
    simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
    as constant propagation. We omit the full set of rules for brevity, but
    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
    \label{fig:semantics_looplets}
\end{figure}

>>>>

@@

    \begin{prooftree}
    \hypo{\langle val, (e, t) \rangle
    \rightarrow val'}
    \infer1[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \llangle g(vals...), t \rrangle}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \hypo{}
        \infer1[$Literal$]{\langle\finchliteral(val), (e, t)\rangle \rightarrow val}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{}
        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \llangle ex, t \rrangle}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t)\rangle}{\rightarrow e(\finchvar(name))}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t)\rangle}{\rightarrow e(\finchindex(name))}}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \hypo{\langle head, s_1 \rangle \rightarrow s_2}
        \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow true}
        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow false}
        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(tns) = tns'}
    \hypo{\langle init, s \rangle \Rightarrow init'}
    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
    \infer4[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle)}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{e(tns) = tns'}
    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \llangle freeze(tns'), t \rrangle)}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = read}
    \hypo{e(tns) = tns'}
    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle)}
    \end{prooftree}

@@

\begin{multline*}
E := [\cdot] | \finchloop(idx, ext, E | body) | \finchblock(E | head, E | tail) | \finchsieve(E | cond, E | body) |\\ 
\finchdeclare(var, E|rhs, E|body) | \finchcall(E|f, E|args...) | \finchaccess(tns, E | idxs...)
\end{multline*}

\begin{prooftree}
    \hypo{e(tns) \mapsto tns'}
    \hypo{\llangle unfurl(tns', ext)\rrangle \mapsto tns''}
    \infer2[$Unfurl$]{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), s\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), s\rangle}
\end{prooftree}
\vspace{6pt}

\begin{prooftree}
    \infer0[$Run$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(run(body), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(body, j...)]), s\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{e(i) = i'}
    \hypo{\llangle seek(i'), t \rrangle \rightarrow t'}
    \infer2[$Lookup$]{\splitfrac{\langle E[\finchaccess(lookup(seek, body), j..., i)], (e, t)\rangle}{\rightarrow \langle E[\finchaccess(body, j...)], (e, t')\rangle}}
\end{prooftree}
\vspace{6pt}

\begin{prooftree}
    \hypo{\llangle cond, t \rrangle \Rightarrow true}
    \infer1[]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(head, i...)], s\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{\llangle cond, t \rrangle \Rightarrow false}
    \infer1[$Switch$]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(tail, i...)], s\rangle}}
\end{prooftree}
\vspace{6pt}

\begin{prooftree}
    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, extent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
\end{prooftree}
\vspace{6pt}

\begin{prooftree}
    \hypo{\llangle preamble, t \rrangle \rightarrow t'}
    \hypo{\langle E[body], (e, t') \rangle \rightarrow (e', t'')}
    \hypo{\llangle epilogue, t'' \rrangle \rightarrow t'''}
    \infer3[$Thunk$]{\langle E[thunk(preamble, body, epilogue)], (e, t')\rangle \rightarrow (e', t''')}
\end{prooftree}
\vspace{6pt}

\begin{prooftree}
    \hypo{\langle \finchloop(i, ext, E[\finchaccess(head, j..., i)]), s\rangle \rightarrow s'}
    \infer1[$Sequence$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(head, tail), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(tail, j..., i)]), s'\rangle}}
\end{prooftree}
\hfill
\begin{prooftree}
    \hypo{\langle node, algebra \rangle \rightarrow node'}
    \infer1[$Simplify$]{\langle E[node], s\rangle \rightarrow \langle E[node'], s\rangle}
\end{prooftree}
\vspace{6pt}

\begin{prooftree}
    \hypo{\llangle seek(a), t \rrangle \rightarrow t'}
    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t')\rangle}}
\end{prooftree}
\vspace{6pt}

\begin{prooftree}
    \hypo{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t) \rangle \rightarrow (e', t')}
    \hypo{\llangle next, t' \rrangle \rightarrow t''}
    \infer2[$StepperNext$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), (e', t'')\rangle}}
\end{prooftree}

    \caption{Looplet evaluation semantics. 
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the host
    language. This means that rules which modify $t$ are running in the host language. Note that $E$ is an evaluation context that applies anywhere in
    the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to
    hoist conditions and subranges out of loops. However, this also means we must specify
    the priority in which we apply looplet rules, which is as follows:
    $Thunk > Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup > Unfurl$
    Many looplets, most notably the thunk looplet, introduce variables into the
    host language environment.  While variables introduced by a looplet may be
    modified by the looplet itself (steppers often increment some state
    variables), we forbid child looplets from modifying any state variables that
    they didn't introduce. This allows us to treat the \finchvalue node as a
    constant.
    %
    Note: the $Simplify$ rule references $algebra$, which is our variable
    defining a set of straightforward simplification rules. These rules include
    simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
    as constant propagation. We omit the full set of rules for brevity, but
    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
    \label{fig:semantics_looplets}
\end{figure}
